---
title: "Assignment 4"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(gapminder)
library(tidyverse)
library(dplyr)
library(ggplot2)

```


***Task 1 - Activity 2***

For this task I will choose four countries and make a tibble with one row per year and columns for life expectancy. The data set I will use is $\texttt{gapminder}$, and the four countries I will look further into are $\texttt{Norway}$, $\texttt{Canada}$, $\texttt{Zimbabwe}$ and $\texttt{Iran}$.

First: Prepare the data set:
```{r}

library(gapminder)


lexp = gapminder %>% 
  filter(country %in% c('Norway', 'Canada', 'Iran', 'Zimbabwe')) %>% 
           select(country, year, lifeExp)
         


```

When the data set is prepared, the table can be made by first reshaping the data so that the correct data is displayed where it is supposed to:

```{r}

lexp_spread = lexp %>% 
  spread(country, lifeExp)
```

Now we can create the table:

```{r}
lexp_spread %>% 
  knitr::kable()
```

This data can be used to generate a scatterplot of the life expectancy of $\texttt{Iran}$ versus the one of $\texttt{Norway}$: 

```{r}
lexp_spread %>% 
  ggplot(aes(Iran, Norway)) + 
  geom_point() + 
  stat_smooth(method = 'gam', formula = y ~poly(x,2), se = FALSE)
```
There is also added a second degree ploynomial line to indicate the trend of the points.
It is interesting to see that Iran has increased its life expectancy in a much higher rate than Norway has over the same time span. 

***Task 2 - Activity #1***


Parts of the $\texttt{gapminder}$ data set could be linked to some other data set, for instance we could look further into the countries and their respective life- cycle savings hypothesis using the built-in data set $\texttt{LifeCycleSavings}$.

```{r}

df = LifeCycleSavings

#Excerpt of the data set
knitr::kable(head(df,10))
  
```

Let us look further into the percent of the per-capita disposable income (dpi) and the percentage of growth rate of dpi (ddpi).
```{r}


new_df = LifeCycleSavings %>% 
  select(dpi, ddpi) %>% 
  rownames_to_column("country")

head(new_df,10)
```

Let us now join the two datasets using both $\texttt{right_join}$ and $\texttt{left_join}$.
The two sets are of different dimensions, with gapminder having 4 columns and 142 rows whilst our new data set $\texttt{new_df}$ has 3 columns and 50 rows.

*Right_join:*
```{r}

rightj_ex =gapminder %>% 
  select(country, continent, year, gdpPercap) %>% 
  
  right_join(new_df,by=c("country"))

#filter by fixed year and display

rightj_ex %>% 
  filter(year == 1977) %>% 
  knitr::kable()
```

Let us try the same using $\texttt{left_join}$:

*Left_join:*
```{r}
leftj_ex =gapminder %>% 
  select(country, continent, year, gdpPercap) %>% 
  
  left_join(new_df,by=c("country"))

#filter by fixed year and display

leftj_ex %>% 
  filter(year == 1977) %>% 
  knitr::kable()
```

We observe that left_join and right_join returns some rows with missing values. This is because the respective functions returns all the rows from one of the data sets so that in this case, right_join would return 50 rows fitted to the $\texttt{new_df}$ and left_join fits the $\texttt{gapminder}$ set: right_join(x,y) would for instance return all rows from y, and the opposite would happen with left_join(x,y). This results in the displayed data loss.

If we want all rows to be displayed, we could use $\texttt{full_join}$:
```{r}
fullj_ex =gapminder %>% 
  select(country, continent, year, gdpPercap) %>% 
  
  full_join(new_df,by=c("country"))

#filter by fixed year and display


fullj_ex %>% 
  filter(year == 1977) %>% 
  knitr::kable()

nrow(fullj_ex)


```

If we want to only display the number of rows that the two data sets have in common, we could use $\texttt{inner_join}$: 

```{r}


innerj_ex =gapminder %>% 
  select(country, continent, year, gdpPercap) %>% 
  
  inner_join(new_df,by=c("country"))

#filter by fixed year and display


innerj_ex %>% 
  filter(year == 1977) %>% 
  knitr::kable()

nrow(innerj_ex)
```

An option for only displaying the data from one of the sets that would match the other completely, we could use $\texttt{semi_join(x,y)}$. This will keep just the columns from x. The difference between semi join and inner join is that semi join never will return duplicate rows of x:

```{r}
semij_ex =gapminder %>% 
  select(country, continent, year, gdpPercap) %>% 
  
  semi_join(new_df,by=c("country"))

#filter by fixed year and display


semij_ex %>% 
  filter(year == 1977) %>% 
  knitr::kable()

nrow(semij_ex)

semij_ex_switch =new_df %>% 
  select(country, dpi,ddpi) %>% 
  
  semi_join(gapminder, by ='country')

semij_ex_switch %>% 
  knitr::kable()



nrow(semij_ex_switch)

```

If we would like to display all rows of one of the data sets that does not match the rows from the other set, we could use $\texttt{anti_join(x,y)}$, returning the rows of x that are not present in y:

```{r}
antij_ex = gapminder %>% 
  select(country, continent, year, gdpPercap) %>% 
  
  anti_join(new_df,by=c("country"))

#filter by fixed year and display


antij_ex %>% 
  filter(year == 1977) %>% 
  knitr::kable()

nrow(antij_ex)



antij_ex_switch =new_df %>% 
  select(country, dpi,ddpi) %>% 
  
  anti_join(gapminder, by ='country')

antij_ex_switch %>% 
  knitr::kable()



nrow(antij_ex_switch)


```

The two last functions are called filtering joins, since they keep cases from the left-hand data set, excluding rows from right-hand data set in different ways. 

***Conclusion***

We have looked further into both reshaping a data set, using different techniques in order to display the desired information, as well as joining two data sets with the help of different functions.
